Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-15 draft.tex
worked on results section, nearly done

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leaves of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
\com{Quelle}

@@

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
%%
Dividing the subjects into four groups using body-size, the number increases to 402 trees.
%%
To make the results cognitive feasible, we have to abstract the results of the classification.
%%------------
\subsubsection{Visualization of Classification Results}
%%
We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
%%
The error rate has to be judged depending on the number of classes of the target variable.
%%
An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
%%
The same error rate is acceptable a target variable with more than two manifestations.
%%
Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
%%
The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
%%
\begin{equation}
error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
\end{equation}
%%
This allows for comparability of error rates between variables.
%%
Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatterplot.
%%
By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%

Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
We map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows to instantly assess the reliability of the result.
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the interactions.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%---------
\subsection{Results}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}

\subsubsection{All Variables}
% None of the pain indicators can be reliably described using the image-based parameter.
None of the pain indicators can be described reliably using the image-based parameter.
%%
The perfect categorization of the \emph{dialysis} parameter was conspicuous.
%%
This is due to the subject distribution among the manifestation.
%%
Only two subjects receive a dialysis.
%%
The good discrimination is not meaningful.
%%
Gender can be discriminated with an error rate of 31,5\% using 7 rules and incorporates only curvature and curvature angle variables.
%%
Surprisingly high is the classification of medication for high blood pressure.
%%
1,058 subjects are classified with an error rate of 23,8\% solely based on coronal mean curvature.
%%
Other features well classified include marital status, foot pain and number of born children.
%%

The dummy variable analysis yields fewer results.
%%
It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
%%
Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
%%
\subsubsection{Gender Groups}
\subsubsection{Age Groups}
\subsubsection{Size Groups}
%%
Many previously described results can be associated to be confounded with subject size.
%%
Differences in the gender analysis are mostly due to the average height difference between males and females.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into height groups potentially highlights classifications not confounded by body-height.
%%

>>>>

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leaves of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose to the data.
\com{Quelle Decision Tree}

@@

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
%%
Dividing the subjects into four groups using body-size, the number increases to 402 trees.
%%
To make the results cognitive feasible, we have to abstract the results of the classification.
%%------------
\subsubsection{Visualization of Classification Results}
%%
We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
%%
The two measures form the axis for a \emph{scatter plot} of the classification results.

\paragraph{The Error Term. }
Normally the error rate for a classification is calculated with $error = \frac{n_{classifiedCorrectly}} {n}$ where $n$ is the number of subjects.
%%
This metric says nothing about the discriminative power of each manifestation.
%%
If for example a variable indicating a disease is negative for 90\% of the subject and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
%%
Therefore the discriminative power of each dimension in the equation
%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
\begin{equation}
error = 1 - \frac{\sum_{m=0}^M \frac{m_{classifiedCorrectly}}{m_{all}}}{number_{dimensions}}
\end{equation}
where $M$ represents each manifestation of a variable.
%%
The error is scaled to denote perfect classification with 0 and 1 is equal to random selection.
% The error rate has to be judged depending on the number of classes of the target variable.
% %%
% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
% %%
% The same error rate is acceptable a target variable with more than two manifestations.
% %%
% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
% %%
% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
% %%
% \begin{equation}
% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
% \end{equation}
%%
This allows for comparability of error rates between variables.

\paragraph{Attribute Mapping}
The scatter plot axis are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
Lets say we want to classify and compare the same variables for different subject groups, for example male and female subjects.
%%
We can support this by color coding the data points according to group affiliation.

Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
We map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows to instantly assess the reliability of the result.
%%
\com{ToDo: Manche Variablen rausgeschmissen, weil sie zu wenig Probanden enthielten, oder Auspraegungen nur duenn besetzt waren!}

\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%---------
\subsection{Results}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.

\subsubsection{All Variables}
%%
The vast majority of parameters can not be described well using the classifier.
%%
This is reflected in the large amount of variables classified with a error rate above $0,6$.
%%
%The first impression of the resulting scatter plots quickly shows a very high amount of variables above an error rate of $0,6$.
%%

% None of the pain indicators can be reliably described using the image-based parameter.
None of the pain indicators can be described reliably using the image-based parameter.
%%
The only variable reliably classified in this group is gender.% with an error rate of 31,5\% using a decision tree of seven rules.
%
It can be discriminated with an error rate of $0,31$ using 7 rules and incorporates only curvature and curvature angle variables.
%
%It mostly incorporates \emph{mean curvature}, \emph{sagittal curvature angle} und sagittal \emph{mean curvature}.
%The perfect categorization of the \emph{dialysis} parameter was conspicuous.
%%
%This is due to the subject distribution among the manifestation.
%%
%Only two subjects receive a dialysis.
%%
%The good discrimination is not meaningful.
%%
Surprisingly high is the classification of medication for high blood pressure.
%%
1,058 subjects are classified with an error rate of $0,47$ solely based on coronal mean curvature.
%%
Almost all subjects who are medicated ($796/1,058$) were correctly classified, the vast majority of non-medicated subjects ($262/1,058$) are false-positive classified.
%%
Therefore the classifier is not as useful as the error-rate would indicate.
%%This is however due to the high amount of people, who are medicated (68\%) and who are almost all correctly class
%%
%Other features well classified include marital status, foot pain and number of born children.
%%
The four body size groups could be characterized with an error rate $0,48$, but the decision tree comprises of 71 rules and imposes overfitting.

The analysis of the dummy variable yields similar results like the blood pressure medication.
%%
Variables, such subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields a error rate below $0,5$.
%The dummy variable analysis yields fewer results.
%%
%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
%%
%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
%%
\subsubsection{Gender Groups}
%%
Classification using groups divided by gender do not produce satisfying results.
%%
Only hypothyroidism could be described for male subjects with an error rate of $0,24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with hypothyroidism, reducing the statistical power of the result.
%%
The dummy variable analysis showed that female subjects of 139-164 cm body height could be discriminated using the mean curvature and curvature angle, with an error of $0,38$.

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0,35$ ($43,8-54$ years of age, 3 decision rules) to $0,27$ ($21-43,8$ years of age, 6 decision rules).
%%
Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between $43,8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0,32$.

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0,4$ and $0,5$.
%%
It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.

\subsubsection{Size Groups}
%%
Many previously described results can be associated to be confounded with subject size.
%%
Differences in the gender analysis are mostly due to the average height difference between males and females.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into height groups potentially highlights classifications not confounded by body-height.
%%

\com{Schmerzparameter mit drin, aber nur mit komplexen bäumen}

\com{- Kinderanzahl: 630 cases, 3,7\% Fehlerrrate, liegt aber daran, dass die meisten Probanden kein Kind haben. Die Einteilung ist aktuell auch unsinnig, das erste Quantil hat (-.0004,0.8] Kinder. Die Probanden werden nicht sauber erkannt.}

\com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B. bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben. Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0. Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}
